Developing a Visual Studio Code Language Support Extension for the Snail Programming Language

### 2023-06-16 paper.tex
feat: draft sec 5.1

This paper will outline the process of building the Snail Language Support VS Code extension.\footnote{This paper is formatted using \LaTeX. Recognizing that using \LaTeX \hspace{1pt} is really hard, the \LaTeX \hspace{1pt} source code for this paper can be found hosted on GitHub: \url{https://github.com/CharlesReinhardt/snail-language-support-paper/}} Specificaly, we will introduce the structure of a VS Code extension that is meant to add support for new programming langauges. We will also discuss how this extension uses VS Code's language server protocol (LSP) to provide realtime error diagnostics \cite{Microsoft_2022a}. We will address how the Snail Language Support extension may be further developed to include debugging support with breakpoints, step in/out behavior, and variable inspection, particularly highlighting the role of VS Code's debug adapter protocol (DAP) \cite{Microsoft_2021}. Finally, we will review good software development practices such as version control and documentation.

@@

\subsection{Theory}

>>>>

This paper will outline the process of building the Snail Language Support VS Code extension.\footnote{This paper is formatted using \LaTeX. Recognizing that using \LaTeX \hspace{1pt} is really hard, the \LaTeX \hspace{1pt} source code for this paper can be found hosted on GitHub: \url{https://github.com/CharlesReinhardt/snail-language-support-paper/}} Specificaly, we will introduce the structure of a VS Code extension that is meant to add support for new programming langauges. We will also discuss how this extension uses VS Code's language server protocol (LSP) to provide realtime error diagnostics \cite{Microsoft_2022a}. We will address how the Snail Language Support extension may be further developed to include debugging support with breakpoints, step in/out behavior, and variable inspection, particularly highlighting the role of VS Code's debug adapter protocol (DAP) \cite{Microsoft_2021a}. Finally, we will review good software development practices such as version control and documentation.

@@

Snail Language Support is also structured to support the implementation of a debugger for snail. To do this, Snail Language Support defines a debug adapter that uses VS Code's Debug Adapter Protocol (DAP). In this section, we will discuss what the DAP is and how Snail Language Support uses it.

\subsection{What is the Debug Adapter Protocol?}

Debugging tools are an integral part of the software development process. Modern debuggers are often implemented with a user-friendly UI to help developers debug their programs \cite{Microsoft_2022b,Matloff_Salzman_2008,Sinha_2017}. Many of these debuggers are built intertwined with their corresponding code editor's UI, which can vary from code editor to code editor. As a result, valuable debugging-logic is intermingled with code-editor-specific UI logic. This makes it nearly impossible to implement a debugger that is capable of communicating with different code editors \cite{Microsoft_2021a}. This problem is illustrated in figure \ref{fig:no-dap-theory}.

Instead of asking a development tool to interface directly with a debugger, suppose a development tool interfaces with a debug adapter that translates a development tool's requests and sends them to a language-specific (but tool-agnostic) debugger. If the communication between the development tool and debug adapter is standardized, a language-specific debug adapter can be reused across multiple development tools.

VS Code's Debug Adapter Protocol (DAP) standardizes the communication between a development tool and debug adapter. As shown in figure \ref{fig:dap-theory}, this allows language-specific debuggers and debug adapters to be reused for multiple development tools.

VS Code's DAP relies on requests and responses very similarly to the LSP \cite{Microsoft_2021b}. A request is a message that a development tool sends to a debug adapter, asking the debugger to perform some debugging task. The debug adapter communicates with the language-specific debugger to perform this operation and sends a corresponding response message to the development tool. The debug adapter \textbf{must} send a corresponding response after recieving a request, even if the response is an empty message. The DAP also defines reverse requests, which function the same as regular requests except that they are sent from the debug adapter to the development tool.

VS Code's DAP also allows a debug adapter to send the development tool events. This allows the debug adapter to update the development tool on a debugging session's state. For example, a debug adapter can send the development tool a \lstinline{Stopped} event when program execution has stopped due to a breakpoint, step request completion, or expression evaluation completion.

\begin{figure}[b]
    \begin{subfigure}[c][][c]{0.45\textwidth}
        \includegraphics[width=\textwidth]{no-dap-theory.png}
        \caption{Communication between a generic debugger and a development tool is incredibly complicated.}
        \label{fig:no-dap-theory}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}[c][][c]{0.45\textwidth}
        \includegraphics[width=\textwidth]{dap-theory.png}
        \caption{With the Debug Adapter Protocol, a development tool can easily communicate with a generic langauge-specific debugger.}
        \label{fig:dap-theory}
    \end{subfigure}
    \caption{Illustrating the benefit of the Debug Adapter Protocol in aiding development tool and debugger communication.}
    \label{fig:dap}
\end{figure}
